Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-11-07 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Fix several minor issues noticed by the reviewer

In every inference rule, (sub-)formulas in the conclusion correspond\footnote{This correspondence could be defined formally, but it would be tedious to do so.} to formulas in the premises (even in the case of quantifier rules, where the formulas differ due to substitution). For any expression occurring both in a formula in the premise and in its corresponding (sub-)formula in the conclusion, the occurrence in the premise is said to be an \emph{ancestor} of the occurrence in the conclusion, and the occurrence in the conclusion is said to be a \emph{descendant} of the occurrence in the premise.

@@

\begin{axiom}
\label{A1}
Either a property or its negation is positive, but not both:
$$
\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))
$$
\end{axiom}

@@

\begin{small}
\begin{prooftree}
		\AXC{$\subproof_2$} \dashedLine
		\UIC{$P(\lambda x. x \neq x)\scott$}    
    		\AXC{$\subproof_1$} \dashedLine
			  \UIC{$P(\lambda x. x = x)\scott$} 
					\AXC{Half of Axiom 1} \dashedLine
					\UIC{$P(\lambda x. x = x) \imp \neg P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
				\BIC{$\neg P(\lambda x. x \neq x)$} \RightLabel{$\neg_E$}
			\BIC{$[\subproof_3] \qquad \bot$}
\end{prooftree}
\end{small}

@@

\begin{axiom}[As in Scott's]
\label{A:A4}
Positive properties are necessarily positive:
$$
\all \varphi.(P(\varphi) \to \Box \; P(\varphi))
$$
\end{axiom}

\begin{definition}[Emended]
\label{A:D2}
An \emph{essence} of an individual is a property possessed by it and necessarily implying those and only those properties necessarily possessed by it:
$$
\ess{\varphi}{x} \equiv \varphi(x) \wedge \all \psi. (\nec \psi(x) \biimp \nec \all x. (\varphi(x) \imp \psi(x)))
$$
\end{definition}

@@

%implication for equivalence in essence: 
%\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A3}} \dashedLine
\UIC{$P(G)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec  \forall y. (G(y)\imp \varphi(y))$}\RightLabel{$\wedge_I$}
\BIC{$P(G)\wedge \nec  \forall y. (G(y)\imp \varphi(y))$}\RightLabel{$\imp_E$}
\AXC{Axiom \ref{A:A2}} 
\BIC{$P(\varphi)$}
\AXC{$\subproof_9$} \dashedLine 
\UIC{$  G(\gamma)$}  \RightLabel{D\ref{A:D1}}
\UIC{$ \all \varphi. (P(\varphi)\biimp \nec \varphi(\gamma))$}  \RightLabel{$\all_E$}
\UIC{$ P(\varphi)\biimp \nec \varphi(\gamma)$}  \RightLabel{$\biimp_E$}
\BIC{$ \nec \varphi(\gamma)$} \RightLabel{$\imp_I^5$}
\UIC{$[\subproof_{12}] \quad \nec  \forall y. (G(y)\imp \varphi(y))\imp \nec \varphi(\gamma)$}
\end{prooftree}
%\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_{11}$ and $\subproof_{12}$} \dashedLine \RightLabel{$\wedge_I$}
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. (G(y)\imp \varphi(y)) $} \RightLabel{$\forall_I$}
\UIC{$\forall \varphi. (\nec \varphi(\gamma)\biimp \nec  \forall y. (G(y)\imp \varphi(y))) $}
\AXC{$ \subproof_{9}$} \dashedLine
\UIC{$G(\gamma)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_9$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\biimp_E$}
\BIC{$\nec E(\gamma)$}\RightLabel{$\nec_E$}
   % \AXC{Axiom T} \dashedLine
    %\UIC{$\nec E(\gamma) \imp E(\gamma)$} \RightLabel{$\imp_E$}
\UIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{A:D3}}
\UIC{$ \all \varphi.(\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)) $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
%\end{tiny}
\end{proof}

@@

\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. (P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(G) \biimp \nec G(\gamma)$} \RightLabel{$\biimp_E$}
\AXC{Axiom \ref{A:A3}} \RightLabel{}
\UIC{$P(G)$} \RightLabel{$\imp_E$}
\BIC{$\nec G(\gamma)$}
\AXC{$ $} \RightLabel{1}
\UIC{$G(\gamma)$} \RightLabel{$\ex_I$}
\UIC{$\ex x.G(x)$}  \RightLabel{$\imp_I^1$}
\UIC{$G(\gamma)\imp \ex x.G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec (G(\gamma)\imp \ex x.G(x))$} \doubleLine  \RightLabel{(Axiom K)}
\UIC{$\nec G(\gamma)\imp \nec \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\nec \ex x.G(x)$}\RightLabel{$\imp_I^2$}
\UIC{$\ex z. G(z)\imp \nec \ex x.G(x)$}
\end{prooftree}
%\end{scriptsize}

@@

\noindent
Proofs of other lemmas, theorems and corollaries remain the same as in the previous section. Theorem \ref{A:T1} is derived from axioms \ref{A:A1} and \ref{A:A2}. Corollary \ref{A:C1} is derived from Axiom \ref{A:A3} and Theorem \ref{A:T1}. Finally, Theorem \ref{A:T3} can be derived from Corollary \ref{A:C1} and Lemma \ref{A:L1}, which, as shown above, depends only on Axiom \ref{A:A3} and Definition \ref{A:D1}. Therefore, confirming H\'ajek's claim, axioms \ref{A:A4} and \ref{A:A5} and definitions \ref{A:D2} and \ref{A:D3} are superfluous: they are not needed to prove Theorem \ref{A:T3}.

@@

\begin{small}
\begin{prooftree}
\AXC{Theorem 2}\dashedLine
\UIC{$\all y.(G(y) \imp \ess{G}{y})$} \RightLabel{D2 }\doubleLine
\UIC{$\all y.(G(y) \imp G(y) \wedge \all \psi. (\nec \psi(y) \biimp \nec \all x. (G(x) \imp \psi(x))))$}
\UIC{$\all y.(G(y) \imp \all \psi. (\nec \psi(y) \biimp \nec \all x. (G(x) \imp \psi(x))))$} \doubleLine
%\UIC{$\all y.(G(y) \imp (\nec A(y) \biimp \nec \all x. (G(x) \imp A(x))))$} \RightLabel{$A$ is constant}
\UIC{$\all y.(G(y) \imp (\nec A \biimp \nec \all x. (G(x) \imp A)))$} \doubleLine
\UIC{$[\subproof_{13}] \qquad \ex y.G(y) \imp (\nec A \biimp \nec \all x. (G(x) \imp A))$}
\end{prooftree}
\end{small}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.G_B(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.G_B(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: G_B(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.G_B(x) $} \RightLabel{$\neg_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg G_B(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  G_B(x) \imp \neg G_B(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(G_B(x) \imp \neg G_B(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(G_B(x) \imp \neg G_B(x)) $}\dottedLine  \RightLabel{D\ref{B:D1}}
\UIC{$P(\neg G_B)$}
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg G_B) \imp \neg P(G_B))$} \RightLabel{$\all_E$}
\UIC{$ P( \neg G_B(x)) \imp \neg P(G_B) $}  \RightLabel{$\imp_E$}
\BIC{$\neg P(G_B) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(G_B)$} \RightLabel{$\neg_E$}
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.G_B(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.G_B(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(G_B) \imp \pos \ex x.G_B(x) $}

@@

\begin{corollary}
\label{B:C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}
\begin{proof}

\begin{prooftree}
\AXC{$ $}\doubleLine
\UIC{$\nec \all x. (G_B(x) \imp G_B(x)) $}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$P_B(G_B) $}\RightLabel{ $\wedge_I$}
\AXC{Theorem \ref{T1}}
\UIC{$ P_B(G_B) \imp \pos \ex x.G_B(x) $} \RightLabel{ $\imp_E$}
\BIC{$\pos \ex x.G_B(x) $}
\end{prooftree}

@@

>>>>

In every inference rule, (sub)formulas in the conclusion correspond\footnote{This correspondence could be defined formally, but it would be tedious to do so.} to formulas in the premises (even in the case of quantifier rules, where the formulas differ due to substitution). For any expression occurring both in a formula in the premise and in its corresponding (sub)formula in the conclusion, the occurrence in the premise is said to be an \emph{ancestor} of the occurrence in the conclusion, and the occurrence in the conclusion is said to be a \emph{descendant} of the occurrence in the premise.

@@

\begin{axiom}
\label{A1}
Either a property or its negation\footnote{Throughout the paper, $\neg \varphi$ is an abbreviation for $\lambda x. \neg \varphi(x)$} is positive, but not both:
$$
\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))
$$
\end{axiom}

@@

\begin{small}
\begin{prooftree}
		\AXC{$\subproof_2$} \dashedLine
		\UIC{$P(\lambda x. x \neq x)\scott$}    
    		\AXC{$\subproof_1$} \dashedLine
			  \UIC{$P(\lambda x. x = x)\scott$} 
					\AXC{Contraposition of Half of Axiom 1} \dashedLine
					\UIC{$P(\lambda x. x = x) \imp \neg P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
				\BIC{$\neg P(\lambda x. x \neq x)$} \RightLabel{$\neg_E$}
			\BIC{$[\subproof_3] \qquad \bot$}
\end{prooftree}
\end{small}

@@

\begin{axiom}[As in Scott's]
\label{A:A4}
Positive properties are necessarily positive:
$$
\all \varphi.(P(\varphi) \to \Box P(\varphi))
$$
\end{axiom}

\begin{definition}[Emended]
\label{A:D2}
An \emph{essence} of an individual is a property necessarily implying those and only those properties necessarily possessed by it:
$$
\ess{\varphi}{x} \equiv \all \psi. (\nec \psi(x) \biimp \nec \all x. (\varphi(x) \imp \psi(x)))
$$
\end{definition}

@@

%implication for equivalence in essence: 
%\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A3}} \dashedLine
\UIC{$P(G)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec  \forall y. (G(y)\imp \varphi(y))$}\RightLabel{$\wedge_I$}
\BIC{$P(G)\wedge \nec  \forall y. (G(y)\imp \varphi(y))$}\RightLabel{$\imp_E$}
\AXC{Axiom \ref{A:A2}} 
\BIC{$P(\varphi)$}
\AXC{$\subproof_9$} \dashedLine 
\UIC{$  G(\gamma)$} \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$ \all \varphi. (P(\varphi)\biimp \nec \varphi(\gamma))$}  \RightLabel{$\all_E$}
\UIC{$ P(\varphi)\biimp \nec \varphi(\gamma)$}  \RightLabel{$\biimp_E$}
\BIC{$ \nec \varphi(\gamma)$} \RightLabel{$\imp_I^5$}
\UIC{$[\subproof_{12}] \quad \nec  \forall y. (G(y)\imp \varphi(y))\imp \nec \varphi(\gamma)$}
\end{prooftree}
%\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_{11}$ and $\subproof_{12}$} \dashedLine \RightLabel{$\wedge_I$}
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. (G(y)\imp \varphi(y)) $} \RightLabel{$\forall_I$}
\UIC{$\forall \varphi. (\nec \varphi(\gamma)\biimp \nec  \forall y. (G(y)\imp \varphi(y))) $}
\dottedLine \RightLabel{D\ref{A:D2}}
\UIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_9$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\biimp_E$}
\BIC{$\nec E(\gamma)$}\RightLabel{$\nec_E$}
   % \AXC{Axiom T} \dashedLine
    %\UIC{$\nec E(\gamma) \imp E(\gamma)$} \RightLabel{$\imp_E$}
\UIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{A:D3}}
\UIC{$ \all \varphi.(\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)) $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
%\end{tiny}
\end{proof}

@@

\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. (P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(G) \biimp \nec G(\gamma)$} \RightLabel{$\biimp_E$}
\AXC{Axiom \ref{A:A3}} \RightLabel{}
\UIC{$P(G)$} \RightLabel{$\biimp_E$}
\BIC{$\nec G(\gamma)$}
\AXC{$ $} \RightLabel{1}
\UIC{$G(\gamma)$} \RightLabel{$\ex_I$}
\UIC{$\ex x.G(x)$}  \RightLabel{$\imp_I^1$}
\UIC{$G(\gamma)\imp \ex x.G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec (G(\gamma)\imp \ex x.G(x))$} \doubleLine  \RightLabel{(Axiom K)}
\UIC{$\nec G(\gamma)\imp \nec \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\nec \ex x.G(x)$}\RightLabel{$\imp_I^2$}
\UIC{$\ex z. G(z)\imp \nec \ex x.G(x)$}
\end{prooftree}
%\end{scriptsize}

@@

\noindent
Proofs of other lemmas, theorems and corollaries remain the same as in the previous section. Theorem \ref{A:T1} is derived from Axioms \ref{A:A1} and \ref{A:A2}. Corollary \ref{A:C1} is derived from Axiom \ref{A:A3} and Theorem \ref{A:T1}. Finally, Theorem \ref{A:T3} can be derived from Corollary \ref{A:C1} and Lemma \ref{A:L1}, which, as shown above, depends only on Axiom \ref{A:A3} and Definition \ref{A:D1}. Therefore, confirming H\'ajek's claim, axioms \ref{A:A4} and \ref{A:A5} and definitions \ref{A:D2} and \ref{A:D3} are superfluous: they are not needed to prove Theorem \ref{A:T3}.

@@

\begin{small}
\begin{prooftree}
\AXC{Theorem 2}\dashedLine
\UIC{$\all y.(G(y) \imp \ess{G}{y})$} \RightLabel{D2 }\doubleLine
\UIC{$\all y.(G(y) \imp \all \psi. (\nec \psi(y) \biimp \nec \all x. (G(x) \imp \psi(x))))$} \doubleLine
%\UIC{$\all y.(G(y) \imp (\nec A(y) \biimp \nec \all x. (G(x) \imp A(x))))$} \RightLabel{$A$ is constant}
\UIC{$\all y.(G(y) \imp (\nec A \biimp \nec \all x. (G(x) \imp A)))$} \doubleLine
\UIC{$[\subproof_{13}] \qquad \ex y.G(y) \imp (\nec A \biimp \nec \all x. (G(x) \imp A))$}
\end{prooftree}
\end{small}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.G_B(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.G_B(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: G_B(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.G_B(x) $} \RightLabel{$\neg_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg G_B(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  G_B(x) \imp \neg G_B(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(G_B(x) \imp \neg G_B(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(G_B(x) \imp \neg G_B(x)) $}\dottedLine  \RightLabel{D\ref{B:D1}}
\UIC{$P(\neg G_B)$}
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg \varphi) \imp \neg P(\varphi))$} \RightLabel{$\all_E$}
\UIC{$ P( \neg G_B(x)) \imp \neg P(G_B) $}  \RightLabel{$\imp_E$}
\BIC{$\neg P(G_B) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(G_B)$} \RightLabel{$\neg_E$}
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.G_B(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.G_B(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(G_B) \imp \pos \ex x.G_B(x) $}

@@

\begin{corollary}
\label{B:C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G_B(x)
$$
\end{corollary}
\begin{proof}

\begin{prooftree}
\AXC{$ $}\doubleLine
\UIC{$\nec \all x. (G_B(x) \imp G_B(x)) $}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$P_B(G_B) $} 
\AXC{Theorem \ref{T1}} \dashedLine
\UIC{$ P_B(G_B) \imp \pos \ex x.G_B(x) $} \RightLabel{ $\imp_E$}
\BIC{$\pos \ex x.G_B(x) $}
\end{prooftree}

@@

\paragraph{Acknowledgements:} This work was supported by an Stipendium of the \"Osterreichische Akademie der Wissenschaften (APART). We thank Christoph Benzm\"uller for collaborations on several related works.
